---
title: "Introduction to ggbrain"
author: "Michael Hallquist"
date: "3 Oct 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggbrain)
```

# Images that will be used in the demo

The package includes a few images that can be used for demonstrating its capabilities. The locations of the images (and the code to load them) are listed below.

```{r}
underlay_3mm <- system.file("extdata", "mni_template_2009c_3mm.nii.gz", package = "ggbrain")
echange_overlay_3mm <- system.file("extdata", "echange_overall_zstat_3mm.nii.gz", package = "ggbrain")
schaefer200_atlas_3mm <- system.file("extdata", "Schaefer_200_7networks_2009c_3mm.nii.gz", package = "ggbrain")
print(c(underlay_3mm, echange_overlay_3mm, schaefer200_atlas_3mm))
```

# Overview of ggbrain

The `ggbrain` package uses the `ggplot2` package to create 2D volume renderings of NIfTI files containing MRI data. It seeks to follow the principles of a graphical grammar in which plots are built from datasets that are mapped onto the visual space via a set of aesthetic mappings. Likewise, `ggbrain` follows the approach of creating layers in which different geometric marks (e.g., square pixels) are overlaid on the plot, such that many images and marks can be combined.

## The hierarchy of a ggbrain plot

Every ggbrain plot is composed of one or more panels that have layers representing different images that have been overlaid. Each panel contains a slice along one of the three axes of the volume: x (Left-Right), y (Anterior-Posterior), or z (Superior-Inferior). Panels can also contain annotations that reflect one or more marks to aid in the interpretation of the plot, as well as region labels in which text is positioned at certain markers within the displayed brain slice.

- Plot
  - Panels (aka Slices)
    - Layers: each pixel in the 2D slice contains a value such as a test statistic from a voxelwise analysis. Pixel values can also be categorical, such as levels of a factor (e.g., lobe).
    - Annotations: custom text or graphical marks overlaid on the panel using the ggplot2 `annotate` function (e.g., coordinate label)
    - Region Labels: a variant of ggplot2 `geom_text`, region labels represent text that should be overlaid at one or more spatial positions across the panel/slice (e.g., anatomical labels)
    
# Building a ggbrain plot layer by layer

One can create an empty ggbrain object using the `ggbrain` function. This is akin to creating an empty ggplot object using `ggplot()`.

```{r}
gg_obj <- ggbrain()
```

Consistent with the `ggplot2` approach, additional elements can be added using the `+` operator, as we elaborate below. The main thing to know at this point is that **you need to add all ggbrain-specific elements before adding typical ggplot components such as `theme()`**.

Also note that the order of the addition operations for images and slices (`add_images` and `add_slices`, respectively) is not important. The example order below -- adding images, slices, layers, and labels -- is conceptually intuitive, but could be reordered without altering the plut. As in `ggplot2`, the order of all geom_* layers *does* matter, such that geom_* layers that are added later are rendered on top of layers added earlier. For example, the `geom_brain()` layer for an underlay image should be added before an overlay.

## Add images

A ggbrain plot must have at least one 3D image stored in NIfTI format. These images are read into ggbrain using the `RNifti` package. All images for a given plot must have the same image dimensions (otherwise, an error will occur). At present, `ggbrain` provides no internal support for interpolation to combine images with different resolutions. We encourage you to use `3dresample` from AFNI for this purpose (see details in the 'Other considerations' section below).

The file names containing relevant images can be passed in as arguments to `ggbrain` or they can be appended using `add_images()` In either, it is important to provide a name for the image that is used internally to refer to the image in contrasts, labels, or other operations. If you do not provide a name, you will see a warning:

```
Warning message:
In private$set_images(images, fill_holes, clean_specks) :
  The images vector does not contain any names. This may lead to weird behaviors downstream if 'underlay' and 'overlay' are requested.
```

```{r}
gg_obj <- gg_obj + add_images(c(underlay = underlay_3mm))
```

Note that the image name of 'underlay' has a special meaning in the package. Specifically, using this name will set a default color scale that is grayscale and will not show the legend of this scale in the plot. You don't have to use 'underlay' but it can make things a bit simpler if you have a conventional plot that shows an anatomical scan underneath functional activations.

You can add many images at once by providing a named vector:

```{r, eval=FALSE}
gg_obj <- gg_obj + add_images(c(underlay = underlay_3mm, echange = echange_overlay_3mm))
```

Or you can add many images in independent `add_images` steps. Note that the order of addition is not important.

```{r, eval=FALSE}
gg_obj <- gg_obj +
  add_images(c(underlay = underlay_3mm)) +
  add_images(c(echange = echange_overlay_3mm))
```

## Adding slices

After adding images to the ggbrain object, the next step is to specify which slices should be displayed on the plot. This is achieved using the `add_slices()` function. Following standard neuroimaging terminology, we can add axial, sagittal, and coronal slices. 

The order of the addition operations, whether within a single `add_slices()` call or in a sequence of calls like `add_slices("x=12") + add_slices(c("y=15", "z=10"))`, determines the order of panels on the rendered plot from upper left to lower right (following the standard `facet_wrap()` conventions in ggplot2). Thus, consider the order in which you wish the panels to appear so that the slice additions follow this order.

### Adding slices by spatial position

The most conventional approach is add slices based on their spatial position along one of three cardinal axes, x/sagittal, y/coronal, or z/axial. You can determine the locations of interest using an interactive image viewer such as afni or fsleyes. Or, as described below, you could add a montage of images, add the coordinate labels on the panels, then choose the coordinates of the slices that are most relevant.

You can add one or more slices using `add_slices()` by passing a character vector of positions.

```{r, eval=FALSE}
gg_obj <- gg_obj + add_slices(c("x = 10", "y = 70", "z = 15"))
```

This will add one sagittal slice (x=10), one coronal (y=70), and one axis (z=15).

### Adding slices by quantile

If you want to use standardized coordinates in terms of percentiles, rather than spatial coordinates, you can append `%` at the end of the `add_slices()` specification. This will find the position of the corresponding slice along the axis between 0% and 100%. These percentiles are computed based on the slices that are not entirely empty (i.e., not all zero).

```{r, eval=FALSE}
gg_obj <- gg_obj + add_slices(c("x = 25%", "y = 50%", "z = 90%"))
```

### Adding a single slice with additional aesthetic details

The `add_slices()` function can add many slices at once, but it does not allow you to alter specific aesthetic elements such as the panel title, border, or axis labels. If you would like finer control over the aesthetics of a single slice, use `add_slice()`. This function supports a number of additional arguments that will alter the appearance of the slice, including:

- `title`: The panel title
- `bg_color`: The background color of the panel
- `text_color`: The color of text on the panel
- `border_color`: The color of a border around the panel
- `border_size`: The size (thickness) of the border around the panel. Default is NULL (none).
- `xlab`: The x axis label of the panel
- `ylab`: The y axis label of the panel
- `theme_custom`: A ggplot2 `theme()` (or list of theme calls) added to this panel that customizes its appearance.

```{r, eval=FALSE}
gg_obj <- gg_obj +
  add_slice("x = 25%", bg_color = "gray25", xlab = "x axis", theme_custom = theme(axis.text.x = element_text(size = 24)))
```

### Adding montages

If you would like a set of slices along one of the image planes to form a montage, you can use the `montage()` convenience function to specify the inputs to `add_slices()`. The arguments are 

- `plane`: "sagittal", "coronal", or "axial" ("x", "y", and "z" also work)
- `n`: the number of slices to render along this plane
- `min`: the lowest quantile to be included (between 0 and 1). Default: 0.1
- `max`: the highest quantile to be included (between 0 and 1). Default: 0.9.
- `min_coord`: the lowest spatial position to be included. Mutually exclusive with quantile approach. Default: NULL.
- `max_coord`: the highest spatial position to be included. Mutually exclusive with quantile approach. Default: NULL.

This code adds 10 axial slices between the 10th and 90th percentiles.

```{r, eval=FALSE}
gg_obj <- gg_obj + add_slices(montage("axial", 10, min = 0.1, max = 0.8))
```

This code adds 5 sagittal slices between the x = -10 and x = 10 (millimeters, in image coordinates).

```{r, eval=FALSE}
gg_obj <- gg_obj + add_slices(montage("sagittal", 5, min_coord = -10, max_coord = 10))
```

## Adding brain overlay layers to a plot

Now that we have defined the images to read and the slices to render, we need to map the numeric values in the slices to the points on the plot. This is achieved by `geom_brain()`, which is essentially an extension of `geom_raster()` from ggplot2. The numeric values in the slices fill squares on the plot, essentially forming 'pixels'.

As with standard ggplot2 objects, the color of these pixels is determined by a `scale_fill_*` scale, allowing you to have fine control over the coloration.

Extending `geom_raster()`, the `geom_brain()` layer supports the following standards.

- `interpolate`: if TRUE, `geom_raster` smooths the layer, rather than rendering discrete squares
- `alpha`: A number between 0 and 1 controlling the alpha transparency of the layer
- `show_legend`: If TRUE (default), a legend mapping the numeric values to a color scale is included in the panel.
- `fill`: A color to be used to fill all non-NA pixels. In ggplot2 terms, this is 'setting' the color of all pixels, as opposed to 'mapping' the color to the numeric value via a scale.
- `mapping`: The aesthetic mapping between the data and the raster layer. In general, the default of `aes(fill=value)` should be retained for parametric layers (i.e., where the color of pixels should map to a continuous scale). But as detailed in a subsequent section, the fill can mapped to label columns in categorical image (e.g., atlas).

Relative to standard ggplot2 conventions, the use of `geom_brain()` requires that you specify details about the color scale when the layer is added. This results from the fact that `ggbrain` objects re-use the 'fill' geometry when layers are overlaid. For example, if I have an anatomical underlay with a grayscale fill and a statistical overlay with a color fill, each layer needs to have its own fill scale (with a `scale_fill_*` object). This is achieved internally using the `ggnewscale` package.

The 'take-home message' is that you need to specify fill scale information as part of the `geom_brain()` call. Here are the key arguments:

- `limits`: a two-element vector that if provided, sets the upper and lower bounds on the scale
- `breaks`: a function to draw the breaks on the scale. The default is `integer_breaks()`, which draws a few integer-valued breaks (usually, 5) between the min and max of the scale.
- `unify_scale`: if TRUE (default), the color scale across panels (slices) will be equated. If FALSE, each panel will apply the color scale independently such that the min and max are slice-dependent.
- `fill_scale`: a `scale_fill_*` object use for mapping the numeric values in the data to the fill channel on the plot. Standard scales such as `scale_fill_viridis_c()` can be used for continuous layers. Note that the package also provides a `scale_fill_bisided()` option that, if added, creates separate color scales for the positive and negative pixels on the plot. This is a conventional approach in displays of fMRI activation maps where, for example, positive modulation is mapped to warm colors (reds) while negative modulation is mapped to cool colors (blues). Note that `scale_fill_bisided()` has arguments of `neg_scale` and `pos_scale` that allow the scales used for negative and positive values, respectively, to be specified directly.

### Defining an image layer

The `geom_brain()` function includes an argument called `definition`. This controls the numeric values that are selected for display on the layer. The simplest definition is to name the image to be displayed. Recall that when you use `add_images()`, you provide names for the images that are used as a reference by `ggbrain`.

For example, if I want to render the underlay and overlay layers as-is, I can use the following approach:

```{r, eval=FALSE}
gg_obj <- gg_obj +
  geom_brain(definition = "underlay", fill_scale = scale_fill_gradient(low = "grey8", high = "grey62"), show_legend = FALSE) +
  geom_brain(definition = "overlay", fill_scale = scale_fill_bisided(), show_legend = TRUE)
```

### Defining a subset layers

Oftentimes, we may wish to modify the display by selecting a subset of values to display. One of the most common applications is thresholding of voxelwise statistics so that small/nonsignificant values are omitted. To achieve this, `geom_brain()` support standard subsetting syntax used in R. Here is an example of thresholding the overlay layer so that the absolute (z) statistic is above 2.5:

```{r, eval=FALSE}
gg_obj <- gg_obj +
  geom_brain(definition = "overlay[abs(overlay) > 2.5]", fill_scale = scale_fill_bisided(), show_legend = TRUE)
```

Note how 'overlay' is treated as an R object that is subset using square brackets. The subsetting expression inside the brackets uses standard R syntax and can reference other images currently in the ggbrain object. For example, you could display overlay values only where the underlay is greater than 10:

```{r, eval=FALSE}
gg_obj <- gg_obj +
  geom_brain(definition = "overlay[underlay > 10]", fill_scale = scale_fill_bisided(), show_legend = TRUE)
```

Finally, note that one can combine subsetting expressions using logical operators `&` and `|` , as in standard R syntax for vector operations. For example, here's how you could include only pixels that are both `abs(overlay) > 2.5` and where `underlay > 10`:

```{r, eval=FALSE}
gg_obj <- gg_obj +
  geom_brain(definition = "overlay[abs(overlay) > 2.5 & underlay > 10]", fill_scale = scale_fill_bisided(), show_legend = TRUE)
```

### Defining a contrast layer

The `ggbrain` package also supports full contrasts in which images can be added, subtracted, multiplied, or divided. For example, one might wish to see the difference between one activation map and another. Of course, such contrasts are often computed in the GLM estimation, but it is often useful to compute certain contrasts after the fact.

For example, what if we have signed and absolute reward prediction error maps from a simple reinforcement learning paradigm? We might be interested to look at areas that are more strongly modulated by signed than absolute PEs.

TODO: finish implementation and make up mind about := as contrast name versus level definition.
Maybe `test := choose(1=a > 2; 2=b > 3)` and `test := conjunction(1=a > 2; 2=b > 3; 3 = a > 4)`
where choose would be a left-to-right evaluation and conjunction compute the first-order values, then determine their combinations.

```{r, eval=FALSE}
gg_obj <- gg_obj +
  geom_brain(definition = "diff := absPE - PE; diff[abs(diff) > 1]", fill_scale = scale_fill_bisided(), show_legend = TRUE)
```

## Adding outlines 

In addition to filling in regions with colors, brain plots often benefit from being able to outline, rather than fill, regions. This is accomplished using `geom_outline()`. This geometry finds the boundaries of distinct clusters in the image/contrast and traces the boundaries with a line of some thickness (controlled by `outline_size`).

For continuous images (e.g., z-statistics), one typically *sets* the outline color using `geom_outline(outline="cyan")` or whatever color is desired. This draws the outlines for non-NA voxels in a single color. For categorical images (e.g., anatomical or functional atlases), it may be useful to map the color of the outlines to a categorical label such as network or region. This is accomplished with `geom_outline(mapping=aes(outline=roi_name))` or whatever column in the data corresponds to the categories one wishes to draw with different colors.


## Adjustments to the appearance of image layers:

      blur_edge = NULL,
      fill_holes = NULL, remove_specks = NULL, trim_threads = NULL)


### Cleaning small 'specks' from image slices

At times, when slicing a given image (esp. functional activations), it is possible that some clusters are very small, yielding small 'specks' on some rendered slices. These are visually unappealing and may merit removal, depending on your tastes. Of course, removing these specks can, at the extreme, misrepresent the data, which is undesirable. This concern notwithstanding, you can use the `clean_specks` argument of `add_images` to specify the voxel threshold used to remove clusters smaller than a certain size. For example, `clean_specks = 30` would remove any clusters smaller than 30 voxels in size from each slice on the rendered image.

```{r, eval=FALSE}
# TODO: make this in geom_brain
gg_obj <- gg_obj + add_images(echange=echange_overlay_3mm)
```

### Filling small holes in image slices

Another aesthetic problem is that at times there will be small holes on the interior of a cluster. These may reflect voxels that fall below a statistical threshold (e.g., voxels that have a z-statistic value of 3.02 in a cluster defined by z > 3.1). These can be filled in using the `fill_holes` argument of `add_images`. This specifies the size of holes (in voxels) that should be filled on the rendered slices. For example, `fill_holes = 30` would fill holes of 30 or fewer voxels on the interior of each cluster. The holes are filled by nearest neighbor imputation in which the nearest non-missing voxels are used to impute missing voxels. In integer-valued/categorical images, the mode of the neighboring voxels is used for imputation.

```{r, eval=FALSE}
# TODO: make this in geom_Brain
gg_obj <- gg_obj + add_images(echange=echange_overlay_3mm)
```




# Labeling regions

```{r, eval=FALSE}
a <- ggbrain(bg_color = "gray60", text_color = "black") +
  add_images(c(underlay = "template_brain.nii.gz")) +
  add_images(c(dan_clust = dan_clust), fill_holes = TRUE, clean_specks = 20, labels = dan_labels) +
  # add_slices(paste0("x = ", seq(10, 90, 10), "%")) +
  add_slices(montage("axial", 10, min = 0.1, max = 0.8)) +
  # add_slice("y=50%", title = "hello", xlab = "testx", border_size = 1, border_color = "blue") +
  geom_brain(definition = "underlay", fill_scale = scale_fill_gradient(low = "grey8", high = "grey62"), show_legend = FALSE) +
  geom_brain(definition = "dan_clust", mapping = aes(fill = label), fill_scale = scale_fill_brewer("DAN 17 label", palette = "Set1"), show_legend = TRUE) +
  geom_outline(definition = "dan_clust", mapping = aes(group = roi_label), show_legend = TRUE, outline = "black") + # fill_scale = scale_fill_discrete("hello"),
  # geom_outline(definition = "dan_clust", show_legend = TRUE, mapping=aes(group=label), outline_size=2, outline="cyan") + #fill_scale = scale_fill_discrete("hello"),

  # geom_outline(definition = "dan_clust", show_legend = TRUE, outline = "cyan") + #fill_scale = scale_fill_discrete("hello"),

  geom_region_label_repel(image = "dan_clust", label_column = "roi_label", min.segment.length = 0, size = 1.5, color = "black", force_pull = 0.2, max.overlaps = Inf) +
  render() + plot_layout(guides = "collect")
```

# Saving ggbrain plots to file

# Aesthetic mappings for 

# Combining ggbrain plots

[Principles of patchwork]

# Other considerations


### Resampling images to have the same resolution

Oftentimes, higher-resolution images (e.g., 1mm) yield better graphics since they potentially provide more detail. But fMRI data are often of lower resolution than T1-weighted images. If you wish to render your fMRI data on a higher-resolution anatomical template in `ggbrain`, you can first use AFNI's `3dresample` command to resample your functional data to match the anatomical template (underlay).

Here is an example of resampling a 3mm statistic image to a 1mm underlay using nearest neighbor interpolation. Note that the image to be resampled should be in the same sterotaxic space as the master/template image.

```
3dresample -input echange_overall_zstat_3mm.nii.gz \
  -prefix echange_overall_zstat_1mm.nii.gz \
  -master mni_template_2009c_1mm.nii.gz \
  -rmode NN
```
